Understanding the Causal Impact of Bikeway Installation on Urban Cycling Adoption

### 2024-12-15 paper/paper.qmd
fixed spelling/grammer

---
title: "Understanding the Causal Impact of Bikeway Installation on Urban Cycling Adoption"
subtitle: "A Difference-in-Differences Analysis Shows Protected Bike Lanes Triple Bike Share Ridership in Toronto, 2017-2023"
author: 
  - Steven Li
thanks: "Code and data are available at: [https://github.com/stevenli-uoft/Toronto_BikeShare_Causality](https://github.com/stevenli-uoft/Toronto_BikeShare_Causality)."
date: today
date-format: long
abstract: "We investigate the causal effects of cycling infrastructure improvements on Bike Share ridership in Toronto between 2017 and 2023. Using a difference-in-differences design with over 7.4 million rides across 1,191 bikeways, we find that treated segments experience an additional 133.1 monthly ride per bikeway, translating to nearly 300,000 additional rides annually, or 40% of all rides on treated bikeways post-improvement. Protected lanes generate the largest gains, more than tripling usage compared to shared roadways, while upgrades to existing bikeways outpace the benefits of new installations. Our findings suggest that strategic implementation of protected infrastructure can substantially increase cycling activity, with implications for Toronto's planned addition of 100 kilometers of cycling infrastructure between 2025-2027. These results provide evidence-based guidance for cities working to encourage sustainable transportation through cycling infrastructure development."
format:
  pdf:
    toc: true
include-before: |
  \newpage
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction
Cycling infrastructure plays an essential role in urban transportation, yet quantifying its impact on cycling adoption remains challenging. While previous research demonstrates that protected bike lanes can reduce cycling injuries by up to 89% [@harris2020lane] and increase ridership by up to 171% on specific routes [@nacto2016], less is known about how different types of infrastructure improvements affect bike share system utilization. This knowledge gap is particularly relevant for Toronto, where the city previously committed to significant cycling infrastructure expansion while simultaneously growing its bike share network from 2,854 stations in 2016 to over 6,850 stations in 2020 [@toronto_bike_share_expansion]. Understanding this relationship is essential as cities increasingly rely on bike share systems to complement public transit and provide sustainable transportation options.

@@

Our analysis indicates that infrastructure improvements significantly increase bike share ridership, with effects varying substantially by implementation approach. While we observe modest initial increases of 32.4 rides per month during the treatment year, post-implementation effects grow substantially to 133.1 additional monthly rides per bikeway. In aggregate, this translates to approximately 300,000 additional annual rides across all 186 treated segments, with improvements directly attributable to about 40% of the 1.48 million total rides recorded on these segments in the two years following changes. This substantial impact is particularly notable given that these improvements represent only 16% of bikeways in our study. Protected bike lanes show particularly strong effects, generating 381.9 additional monthly rides post-implementation - more than twice the effect of on-road lanes (169.2 rides) and triple that of shared roadways (111.7 rides). Notably, these gains occurred despite treated bikeways having higher initial ridership at 209.03 monthly rides, compared to control bikeways 85.59 monthly rides pre-treatment. This suggests infrastructure improvements can substantially increase cycling activity even on popular routes.

@@

## Infrastucture Development and Policy Implications
Our findings have substantial implications for cycling infrastructure development strategies. The strong performance of protected lanes, combined with the effectiveness of upgrades, suggests a coupled approach:

@@

## Implementation Strategy
### Sample Size and Target Audience
Our survey will be targeted towards: current and former (In-active user for over 6 months) Bike Share members, and non-bike share users living within 200 meters of Toronto Bike Share Stations. To ensure our surveys statistical reliability and stays within our budget, it will aim to target 3,000 respondents:

@@

## Tradeoffs and Limitations
Our survey methodology faces several important limitations that could affect the validity of our results. First, self-selection and non-response bias may skew our sample toward respondents who are already interested in cycling or have strong opinions about infrastructure, while our incentive structure might attract particular demographic groups, potentially missing key segments of the population. Our reliance on hypothetical scenarios and stated preferences introduces additional uncertainty, as actual behavior may differ significantly from reported intentions, especially regarding route selection and infrastructure usage. Despite careful design to minimize social desirability bias, respondents may still feel pressure to express support for sustainable transportation initiatives or understate safety concerns. Furthermore, temporal limitations of our three-month follow-up period may inadequately capture seasonal variations in cycling behavior and long-term adaptation to infrastructure changes, while the results may not generalize well to other urban contexts or represent the full diversity of potential users. These limitations suggest that while our survey can provide imporant findings into the relationship between infrastructure and ridership, findings should be interpreted alongside other data sources and methodologies.

>>>>

---
title: "Understanding the Causal Impact of Bikeway Installation on Urban Cycling Adoption"
subtitle: "A Difference-in-Differences Analysis Shows Protected Bike Lanes Triple Bike Share Ridership in Toronto, 2017-2023"
author: 
  - Steven Li
thanks: "Code and data are available at: [https://github.com/stevenli-uoft/Toronto_BikeShare_Causality](https://github.com/stevenli-uoft/Toronto_BikeShare_Causality)."
date: today
date-format: long
abstract: "We investigate the causal effects of cycling infrastructure improvements on Bike Share ridership in Toronto between 2017 and 2023. Using a difference-in-differences design with over 7.4 million rides across 1,191 bikeways, we find that treated segments experience 133.1 additional monthly ride per bikeway, translating to nearly 300,000 additional rides annually, or 40% of all rides on treated bikeways post-improvement. Protected lanes generate the largest gains, more than tripling usage compared to shared roadways, while upgrades to existing bikeways outpace the benefits of new installations. Our findings suggest that strategic implementation of protected infrastructure can substantially increase cycling activity, with implications for Toronto's planned addition of 100 kilometers of cycling infrastructure between 2025-2027. These results provide evidence-based guidance for cities working to encourage sustainable transportation through cycling infrastructure development."
format:
  pdf:
    toc: true
include-before: |
  \newpage
number-sections: true
bibliography: references.bib
---

@@

\newpage
# Introduction
Cycling infrastructure plays an essential role in urban transportation, yet quantifying its impact on cycling adoption remains challenging. While previous research indicates that protected bike lanes may reduce cycling injuries by as much as 89% [@harris2020lane] and increase ridership by up to 171% on specific routes [@nacto2016], less is known about how different types of infrastructure improvements affect bike share system utilization. This knowledge gap is particularly relevant for Toronto, where the city previously committed to significant cycling infrastructure expansion while simultaneously growing its bike share network from 2,854 stations in 2016 to over 6,850 stations in 2020 [@toronto_bike_share_expansion]. Understanding this relationship is essential as cities increasingly rely on bike share systems to complement public transit and provide sustainable transportation options.

@@

Our analysis indicates that infrastructure improvements significantly increase bike share ridership, with effects varying substantially by implementation approach. While we observe modest initial increases of 32.4 rides per month during the treatment year, post-implementation effects grow substantially to 133.1 additional monthly rides per bikeway. In aggregate, this translates to approximately 300,000 additional annual rides across all 186 treated segments, with improvements directly attributable to about 40% of the 1.48 million total rides recorded on these segments in the two years following changes. This substantial impact is particularly notable given that these improvements represent only 16% of bikeways in our study. Protected bike lanes show particularly strong effects, generating 381.9 additional monthly rides post-implementation - more than twice the effect of on-road lanes (169.2 rides) and triple that of shared roadways (111.7 rides). Notably, these gains occurred even though treated bikeways started with higher initial ridership (209.03 monthly rides) than control bikeways (85.59 monthly rides) pre-treatment.. This suggests infrastructure improvements can substantially increase cycling activity even on popular routes.

@@

## Infrastructure Development and Policy Implications
Our findings have substantial implications for cycling infrastructure development strategies. The strong performance of protected lanes, combined with the effectiveness of upgrades, suggests a coupled approach:

@@

## Implementation Strategy
### Sample Size and Target Audience
Our survey will be targeted towards: current and former (inactive user for over 6 months) Bike Share members, and non-bike share users living within 200 meters of Toronto Bike Share Stations. To ensure our surveys statistical reliability and stays within our budget, it will aim to target 3,000 respondents:

@@

## Tradeoffs and Limitations
Our survey methodology faces several important limitations that could affect the validity of our results. First, self-selection and non-response bias may skew our sample toward respondents who are already interested in cycling or have strong opinions about infrastructure, while our incentive structure might attract particular demographic groups, potentially missing key segments of the population. Our reliance on hypothetical scenarios and stated preferences introduces additional uncertainty, as actual behavior may differ significantly from reported intentions, especially regarding route selection and infrastructure usage. Despite careful design to minimize social desirability bias, respondents may still feel pressure to express support for sustainable transportation initiatives or understate safety concerns. Furthermore, temporal limitations of our three-month follow-up period may inadequately capture seasonal variations in cycling behavior and long-term adaptation to infrastructure changes, while the results may not generalize well to other urban contexts or represent the full diversity of potential users. These limitations suggest that while our survey can provide important findings into the relationship between infrastructure and ridership, findings should be interpreted alongside other data sources and methodologies.
